---
        title: "Reproducible Research - Activity Analysis Assignment"
author: "Lynda Young"
date: "24 April 2017"
output: 
        html_document: 
        keep_md: yes
---
        
        ```{r setup, include=FALSE, fig.height = 4, fig.width=7}
knitr::opts_chunk$set(echo = TRUE)
```

# Analyis of Activity Data

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and includes the number of steps taken in 5 minute intervals each day.

The steps/analysis performed are as follows:
        
        1. Obtaining and examining the data
2. A frequency plot of the daily total step count (includes the median and mean values)
3. A time-series plot of the average daily step activity by 5 minute time intervals (includes identification of the interval with the highest average steps)
4. Imputation of missing step values and repeating step 2. on corrected data
5. Comparison of average daily step activity by intervals for weekends vs weekdays


##1: Obtain and pre-process data



```{r data }
library(downloader)

# obtain and read in data
setwd("C:/Users/Lynda Young/Documents/Coursera/Reproducible Research/Activity_Analysis_Assignment/RepData_PeerAssessment1")
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileUrl,destfile = "activity.zip")
unzip("./activity.zip", exdir = ".")
activity_data <- read.csv("activity.csv")

# investigate   
summary(activity_data)

# change date from factor to date format
activity_data$date <- as.Date(activity_data$date, "%Y-%m-%d")

```